---
format: gfm
---

This is directly extracted from Julia's original code.

```{r}
rm(list = ls())

###Julia Biesiada
### Title: Data Wrangling - Teams Overviews

## -- 0. Data Loading-----------------------------------------------------------
library(tidyverse)
library(ggplot2)
ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")

# -- 1. Filter out all-star games once -----------------------------------------
ufa_clean <- ufa_throws |>
  filter(home_teamID != "allstars1",
         away_teamID != "allstars1")

# -- 2. Home goals scored / conceded per game, then summed ---------------------
home_score <- ufa_clean |>
  group_by(home_teamID, gameID) |>
  summarise(
    goals_scored   = max(home_team_score, na.rm = TRUE),
    goals_conceded = max(away_team_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(home_teamID) |>
  summarise(
    total_goals_scored   = sum(goals_scored),
    total_goals_conceded = sum(goals_conceded),
    .groups = "drop"
  ) |>
  rename(team = home_teamID)

# -- 3. Away goals scored / conceded per game, then summed----------------------
away_score <- ufa_clean |>
  group_by(away_teamID, gameID) |>
  summarise(
    goals_scored   = max(away_team_score, na.rm = TRUE),
    goals_conceded = max(home_team_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(away_teamID) |>
  summarise(
    total_goals_scored   = sum(goals_scored),
    total_goals_conceded = sum(goals_conceded),
    .groups = "drop"
  ) |>
  rename(team = away_teamID)

# -- 4. Combined goals (home + away) -------------------------------------------
combined_goals <- bind_rows(home_score, away_score) |>
  group_by(team) |>
  summarise(
    total_goals_scored   = sum(total_goals_scored),
    total_goals_conceded = sum(total_goals_conceded),
    .groups = "drop"
  )
# -- 5. Home/Away split stats --------------------------------------------------
split_stats <- ufa_clean |>
  mutate(
    team     = if_else(is_home_team == TRUE, home_teamID, away_teamID),
    location = if_else(is_home_team == TRUE, "home", "away"),
    win      = (is_home_team == TRUE  & home_team_win == 1) |
      (is_home_team == FALSE & home_team_win == 0)
  ) |>
  group_by(team, location) |>
  summarise(
    games     = n_distinct(gameID),
    wins      = n_distinct(gameID[win]),
    losses    = games - wins,
    throws    = sum(turnover == 0 & goal == 0, na.rm = TRUE),
    turnovers = sum(turnover == 1, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  pivot_wider(
    names_from  = location,
    values_from = c(games, wins, losses, throws, turnovers),
    names_glue  = "{location}_{.value}"          # e.g. home_games, away_wins
  )

# -- 6. Core team stats --------------------------------------------------------
team_stats_core <- ufa_clean |>
  mutate(
    team = if_else(is_home_team == TRUE, home_teamID, away_teamID),
    win  = (is_home_team == TRUE  & home_team_win == 1) |
      (is_home_team == FALSE & home_team_win == 0)
  ) |>
  group_by(team) |>
  summarise(
    total_games     = n_distinct(gameID),
    total_wins      = n_distinct(gameID[win]),
    total_losses    = total_games - total_wins,
    total_throws    = sum(turnover == 0 & goal == 0, na.rm = TRUE), #max posession_throw, possesionnumber 1
    total_goals     = sum(goal == 1,     na.rm = TRUE), #check this 
    total_turnovers = sum(turnover == 1, na.rm = TRUE),
    .groups = "drop"
  )

team_stats <- team_stats_core |>
  left_join(combined_goals, by = "team") |>
  left_join(split_stats,    by = "team") |>
  mutate(
    # Ratio: goals scored per throw attempt
    goal_ratio          = round(total_goals / (total_throws + total_goals + total_turnovers), 3),
    
    # Ratio: turnovers per total possession attempts
    turnover_ratio      = round(total_turnovers / (total_throws + total_goals + total_turnovers), 3),
    
    # Throws per goal: How many throws take to score a goal?
    throws_per_goal = round(total_throws / total_goals,1),
    
    # +/- box score: goals scored minus goals conceded across all games
    plus_minus          = total_goals_scored - total_goals_conceded,
    
    # Win percentage
    win_pct             = round(total_wins / total_games,3)
  ) |>
  # Tidy column order
  select(
    team,
    total_games, total_wins, total_losses, win_pct,
    total_throws,throws_per_goal, total_goals, total_turnovers,
    total_goals_scored, total_goals_conceded, plus_minus,
    goal_ratio, turnover_ratio,
    home_games, home_wins, home_losses, home_throws, home_turnovers,
    away_games, away_wins, away_losses, away_throws, away_turnovers
  )

team_stats|>
  mutate(throws_per_goal = total_throws / total_goals)|>
  select(throws_per_goal)

# 7.Adding Tactical Features ---------------------------------------------------

tactics_features <- ufa_clean|>
  mutate(team = if_else(is_home_team == TRUE, home_teamID, away_teamID)) |>
  group_by(team) |>
  summarise( 
    ## Offensive Features
    # Average throw distance on ALL throws
    avg_throw_distance = round(mean(throw_distance, na.rm = TRUE),2),
    # Average throw distance specifically on GOALS
    avg_goal_distance = round(mean(throw_distance[goal == 1], na.rm = TRUE),2),
    # Average throw distance on TURNOVERS 
    avg_turnover_distance = round(mean(throw_distance[turnover == 1], na.rm = TRUE),2),
    # Average throw angle
    avg_throw_angle = round(mean(abs(throw_angle), na.rm = TRUE),3),
    # How often they attempt long throws -> (I assume long throw is >20)
    long_throw_rate = round(sum(throw_distance >= 35, na.rm = TRUE) / n(),3),
    # Long Goal Rate 
    long_goal_rate = round(sum(throw_distance >= 35 & goal == 1,na.rm = TRUE) /
                             sum(goal == 1, na.rm = TRUE), 3),
    
      
    ## Defensive Features
    # Short throw rate — teams that play it safe
    short_throw_rate = round(sum(throw_distance <= 10, na.rm = TRUE) / n(),3),
    # Short Goal Rate 
    short_goal_rate = round(sum(throw_distance <= 10 & goal == 1,na.rm = TRUE) /
                             sum(goal == 1, na.rm = TRUE), 3),
    
    #Medium throw rate 
    medium_throw_rate = round(sum(throw_distance > 10 & throw_distance < 35, na.rm = TRUE) / n(),3),
    # Medium Goal Rate 
    medium_goal_rate = round(sum(throw_distance > 10 & throw_distance < 35 & goal == 1,na.rm = TRUE) /
                             sum(goal == 1, na.rm = TRUE), 3),
    
    #What is the long shot turnover rate?
    long_throw_turnover_rate = round(sum(throw_distance >= 35 & turnover == 1, na.rm = TRUE) /
                                       sum(throw_distance >=35,            na.rm = TRUE), 3),
    # What is the medium shot turnover rate?
    medium_throw_turnover_rate = round(sum(throw_distance > 10 & throw_distance < 20 & turnover == 1, na.rm = TRUE) /
                                        sum(throw_distance > 10 & throw_distance < 20,na.rm = TRUE), 3),
    #What is the short shot turnover rate?
    short_throw_turnover_rate = round(sum(throw_distance <= 10 & turnover == 1, na.rm = TRUE) /
                                      sum(throw_distance >= 10,            na.rm = TRUE), 3),
    
    ## Game Momentum Features (TBD - Add with time stamp and also with the num_possession -> sequence)
    # Goals scored per quarter - who starts strong vs finishes strong?
    goals_q1 = sum(goal == 1 & game_quarter == 1, na.rm = TRUE),
    goals_q2 = sum(goal == 1 & game_quarter == 2, na.rm = TRUE),
    goals_q3 = sum(goal == 1 & game_quarter == 3, na.rm = TRUE),
    goals_q4 = sum(goal == 1 & game_quarter == 4, na.rm = TRUE),
    goals_q5 = sum(goal == 1 & game_quarter == 5, na.rm = TRUE),
    # Turnovers per quarter - when do they lose concentration?
    turnovers_q1 = sum(turnover == 1 & game_quarter == 1, na.rm = TRUE),
    turnovers_q2 = sum(turnover == 1 & game_quarter == 2, na.rm = TRUE),
    turnovers_q3 = sum(turnover == 1 & game_quarter == 3, na.rm = TRUE),
    turnovers_q4 = sum(turnover == 1 & game_quarter == 4, na.rm = TRUE),
    turnover_q5 = sum(turnover == 1 & game_quarter == 5, na.rm = TRUE),
    # Power quarter — which quarter do they score most?
    power_quarter = case_when(
      goals_q1 == pmax(goals_q1, goals_q2, goals_q3, goals_q4) ~ "Q1",
      goals_q2 == pmax(goals_q1, goals_q2, goals_q3, goals_q4) ~ "Q2",
      goals_q3 == pmax(goals_q1, goals_q2, goals_q3, goals_q4) ~ "Q3",
      goals_q4 == pmax(goals_q1, goals_q2, goals_q3, goals_q4) ~ "Q4"
    ),
    
    # Turnover quarter - which quarter do they turnover the most?
    turnovers_quarter = case_when(
      turnovers_q1 == pmax(turnovers_q1, turnovers_q2, turnovers_q3, turnovers_q4) ~ "Q1",
      turnovers_q2 == pmax(turnovers_q1, turnovers_q2, turnovers_q3, turnovers_q4) ~ "Q2",
      turnovers_q3 == pmax(turnovers_q1, turnovers_q2, turnovers_q3, turnovers_q4) ~ "Q3",
      turnovers_q4 == pmax(turnovers_q1, turnovers_q2, turnovers_q3, turnovers_q4) ~ "Q4"
    ))

# 8.Combining Team Stats and Tactical Features ---------------------------------

full_team_stats <- team_stats |>
  left_join(tactics_features, by = "team")

# 9. Visualization for Teams Overview ------------------------------------------

#a) Teams Overview for Total Losses and Total Wins (make it better)

full_team_stats |>
  mutate(team = fct_reorder(team, total_wins, .desc = TRUE))|>
  pivot_longer(cols = c(total_wins, total_losses),
               names_to = "outcome",
               values_to = "count") |>
  ggplot(aes(x = team, y = count, fill = outcome)) +
  geom_col(position = "stack") +
  scale_y_continuous(breaks = seq(0,55, by = 5))+
  scale_fill_manual(values = c("total_wins" = "lightgreen", "total_losses" = "salmon"),
                    labels = c("total_wins" = "Total Wins", "total_losses" = "Total Losses"))+
  labs(title = " Teams Overview",y = "Total Games", x = "Team", fill = "Outcome")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))

#b) 
# Scatter Plot - team patterns 
#left top - elite (score the most and has the least turnovers)
#right top - risky (score a lot but also struggle with turnovers)
#left bottom - safe (can't score but do not make turnovers)
#riht bottom - struggling (hard to score and make a lot of turnovers)
ggplot(full_team_stats, aes(
  x = turnover_ratio,
  y = goal_ratio,
  size = win_pct,
  label = team))+
  geom_point(alpha = 0.8)+
  geom_text(vjust = -1, size = 3) +
  coord_fixed()+
  theme_minimal()


#c) Heat map for the Quarters -> teams and goals by quater 

#d) Heat map for thr Turnovers -> teams and goals by quarter (maybe cutting by teams who played > n games)

#e) Throw Profile per team by creating a bar stack of the short,medium, long throw -> tune the range of the distance

#f) Goal difference (column +-) - like in the pdf - ranking (lolipop)

#g) Home vs Away Win%  -> column to see where teams have better performance

#h) Correaltion map
```

Julia's dataset (take as given):

```{r}
full_team_stats
```

#### 1. Adding other useful metrics to the `full_team_stats` table

Another useful article to check out: https://ultiworld.com/2016/10/25/basic-team-analysis-using-four-metrics/

Note that our current UFA dataset does not divide players into offensive and defensive lines.

```{r}
more_stats <- ufa_clean |>
  mutate(team = case_when(
      is_home_team == TRUE ~ home_teamID,
      is_home_team == FALSE ~ away_teamID
      ),
      throw_path_type = case_when(
        (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
        (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
        TRUE ~ "cross_throw"
        ),
      attacking_phaseID = consecutive_id(gameID, game_quarter, is_home_team,
                                         possession_num, home_team_score,
                                         away_team_score))

more_stats
```

```{r}
# Create different columns to count the right wing, left wing, and cross throws by team
# The throw counts and throw proportions have filtered out the turnovers

throw_positions <- more_stats |>
  filter(turnover == 0) |>
  select(team, throw_path_type) |>
  group_by(team, throw_path_type) |>
  summarize(throw_count = n()) |>
  ungroup() |>
  pivot_wider(names_from = "throw_path_type", values_from = "throw_count") |>
  mutate(total_throw_count = cross_throw + left_wing + right_wing,
         cross_throw_rate = cross_throw / total_throw_count,
         left_wing_rate = left_wing / total_throw_count,
         right_wing_rate = right_wing / total_throw_count,
         dominant_wing = case_when(
           cross_throw == pmax(cross_throw, left_wing, right_wing) ~ "cross",
           left_wing == pmax(cross_throw, left_wing, right_wing) ~ "left",
           right_wing == pmax(cross_throw, left_wing, right_wing) ~ "right"
         )) |>
  mutate(dominant_wing = as.factor(dominant_wing)) |>
  select(team, cross_throw_rate, left_wing_rate, right_wing_rate, dominant_wing)

throw_positions

# Metric of interest: cross_throw_rate, left_wing_rate, right_wing_rate,
# dominant_wing
```

```{r}
# Metric of interest: completion rate for cross throws, left wing, and right wing

throw_completion_wing <- more_stats |>
  select(team, throw_path_type, turnover) |>
  group_by(team, throw_path_type) |>
  summarize(completion_rate = 1 - mean(turnover),
            .groups = "drop") |>
  pivot_wider(names_from = "throw_path_type",
              values_from = "completion_rate") |>
  rename(cross_throw_completion = cross_throw,
         left_wing_completion = left_wing,
         right_wing_completion = right_wing)

throw_completion_wing
```

```{r}
# Metric of interest: conversion rate (number of goals / number of possessions)

attack_outcome_indicators <- more_stats |>
  select(attacking_phaseID, goal, turnover) |>
  group_by(attacking_phaseID) |>
  summarize(sum_goal = sum(goal),
            sum_turnover = sum(turnover)) |>
  ungroup() |>
  mutate(attack_outcome = case_when(
    sum_goal == 1 ~ 1,
    sum_goal == 0 & sum_turnover >= 1 ~ 0,
    sum_goal == 0 & sum_turnover == 0 ~ -1
  )) |>
  select(attacking_phaseID, attack_outcome)

attack_outcome_indicators
```

```{r}
more_stats <- more_stats |>
  left_join(attack_outcome_indicators, join_by(attacking_phaseID)) |>
  mutate(perfect_conversion = ifelse(
    possession_num == 1 & attack_outcome == 1, 1, 0
    ))

more_stats
```

```{r}
# Dropping duplicated data in R:
# https://www.datanovia.com/en/lessons/identify-and-remove-duplicate-data-in-r/

conversion_rate_table <- more_stats |>
  select(team, attacking_phaseID, attack_outcome) |>
  filter(!duplicated(attacking_phaseID)) |>
  select(-attacking_phaseID) |>
  
  # Different from my original data, I will convert attack_outcome to 0
  # if the attacking phase is incomplete. This ensures that attack_outcome is
  # a binary variable so we can summarize the conversion rate easier with mean()
  
  mutate(attack_outcome = ifelse(
    attack_outcome == -1, 0, attack_outcome
  )) |>
  group_by(team) |>
  summarize(conversion_rate = mean(attack_outcome),
            .groups = "drop")

conversion_rate_table
```

```{r}
# Metric of interest: Perfect conversion rate
# This is also similar to conversion rate but needs no prior turnover to count

perfect_conversion_table <- more_stats |>
  select(team, attacking_phaseID, perfect_conversion) |>
  filter(!duplicated(attacking_phaseID)) |>
  select(-attacking_phaseID) |>
  group_by(team) |>
  summarize(perfect_conversion_rate = mean(perfect_conversion),
            .groups = "drop")

perfect_conversion_table
```

Note that this `perfect_conversion_rate` is divided by the total number of attacking phases (possessions) that a team has had.

Putting all together:

```{r}
full_team_stats_updated <- full_team_stats |>
  left_join(throw_positions, join_by(team)) |>
  left_join(throw_completion_wing, join_by(team)) |>
  left_join(conversion_rate_table, join_by(team)) |>
  left_join(perfect_conversion_table, join_by(team))
```

```{r}
full_team_stats_updated
```

```{r}
# Write the resulting dataframe into a CSV file for reference
write.csv(full_team_stats_updated, file = "full_team_stats.csv",
          row.names = FALSE)
```

#### 2. Explore some relationships between two closely related variables with scatter plots

Comparing goal ratio and turnover ratio, building on Julia's scatter plot but with quadrants:

```{r}
# https://stackoverflow.com/questions/17521438/geom-rect-and-alpha-does-this-work-with-hard-coded-values

more_full_team_stats <- full_team_stats_updated |>
  mutate(goal_ratio_demean = scale(goal_ratio,
                                   center = TRUE, scale = FALSE),
         turnover_ratio_demean = scale(turnover_ratio,
                                       center = TRUE, scale = FALSE),
         conversion_rate_demean = scale(conversion_rate,
                                        center = TRUE, scale = FALSE),
         perfect_conversion_rate_demean = scale(perfect_conversion_rate,
                                                center = TRUE, scale = FALSE))

more_full_team_stats |>
  ggplot() +
  
  # High goal ratio, high turnover ratio
  annotate("rect", xmin = 0, ymin = 0,
           xmax = max(more_full_team_stats$turnover_ratio_demean),
           ymax = max(more_full_team_stats$goal_ratio_demean),
           fill = "yellow", alpha = 0.3) +
  
  # Low goal ratio, low turnover ratio
  annotate("rect", xmax = 0, ymax = 0,
           xmin = min(more_full_team_stats$turnover_ratio_demean),
           ymin = min(more_full_team_stats$goal_ratio_demean),
           fill = "yellow", alpha = 0.3) +
  
  # High goal ratio, low turnover ratio
  annotate("rect", xmax = 0, ymin = 0,
           xmin = min(more_full_team_stats$turnover_ratio_demean),
           ymax = max(more_full_team_stats$goal_ratio_demean),
           fill = "green", alpha = 0.3) +
  
  # High turnover ratio, low goal ratio
  annotate("rect", xmin = 0, ymax = 0,
           xmax = max(more_full_team_stats$turnover_ratio_demean),
           ymin = min(more_full_team_stats$goal_ratio_demean),
           fill = "red", alpha = 0.3) +
  
  # Add points and text to the plot
  geom_point(aes(x = turnover_ratio_demean, y = goal_ratio_demean,
                 size = win_pct), alpha = 0.8) +
  geom_text(aes(x = turnover_ratio_demean, y = goal_ratio_demean,
                label = team), vjust = -1, size = 3) +
  
  geom_hline(yintercept = 0, color = "gray30", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray30", linetype = "dashed") +
  coord_fixed() + theme_minimal()
```

Plotting conversion rate and perfect conversion rate:

```{r}
more_full_team_stats |>
  ggplot(aes(x = conversion_rate,
             y = perfect_conversion_rate / conversion_rate)) +
  
  # Add points and text to the plot
  geom_point(aes(size = win_pct), alpha = 0.8) +
  geom_text(aes(label = team), vjust = -1, size = 3) +
  
  # Add a regression line to the plot (very closely related!)
  geom_smooth(method = "lm") +
  
  coord_fixed() + theme_minimal()
```

Plotting cross throw rate and cross throw completion, colored by dominant wing:

```{r}
more_full_team_stats |>
  ggplot(aes(x = cross_throw_rate, y = cross_throw_completion,
             color = dominant_wing)) +
  
  # Add points and text to the plot
  geom_point(aes(size = win_pct), alpha = 0.8) +
  geom_text(aes(label = team), vjust = -1, size = 3) +
  
  # Add a trend line to the plot
  geom_smooth(method = "lm", alpha = 0.5) +
  
  coord_fixed() + theme_minimal()
```

#### 3. How often does each team use each type of throw?

```{r}
full_team_stats_updated |>
  select(team, cross_throw_rate, left_wing_rate, right_wing_rate) |>
  pivot_longer(cross_throw_rate:right_wing_rate,
               names_to = "throw_type", values_to = "proportion") |>
  ggplot(aes(x = team, y = proportion, fill = throw_type)) +
  geom_col()
```

Bonus: Conversion rate and perfect conversion rate based on dominant wing:

```{r}
# fct_reorder and text rotation syntax used from Lecture 3 materials

full_team_stats_updated |>
  select(team, conversion_rate, perfect_conversion_rate, dominant_wing) |>
  # Arrange the teams by descending conversion rate in the plot output
  mutate(team = fct_reorder(team, conversion_rate, .desc = TRUE)) |>
  pivot_longer(conversion_rate:perfect_conversion_rate,
               names_to = "conversion_metric", values_to = "rate") |>
  ggplot(aes(x = team, y = rate, fill = conversion_metric,
             color = dominant_wing)) +
  geom_col(position = "dodge") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

#### Future case study: Empires vs. Mechanix - what makes a team good?

```{r}
# I have actually filtered for more teams than available to see which comparisons are available
full_team_stats_updated |>
  filter(team %in% c("empire", "mechanix")) |>
  arrange(desc(win_pct))
```


#### 3. Power and turnover quarters (by the number of teams)

```{r}
quarter_by_teams <- full_team_stats |>
  count(power_quarter) |>
  left_join(full_team_stats |> count(turnovers_quarter),
             join_by(power_quarter == turnovers_quarter),
             suffix = c("_power", "_turnovers")) |>
  tidyr::replace_na(list(n_power = 0, n_turnovers = 0)) |>
  pivot_longer(n_power:n_turnovers, names_to = "stat", values_to = "value")

quarter_by_teams

quarter_by_teams |>
  ggplot(aes(x = power_quarter, y = value, color = stat)) +
  geom_point()
```

#### Additional: Checking Julia's original code with a subset of the data:

```{r}
library(tidyverse)
library(ggplot2)
ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")
```

```{r}
# Filter out all-star games
ufa_clean <- ufa_throws |>
  filter(!(home_teamID %in% c("allstars1", "allstars2")),
         !(away_teamID %in% c("allstars1", "allstars2")))

ufa_clean
```

```{r}
test_data <- ufa_clean |>
  filter(gameID %in% c("2023-06-24-DC-BOS", "2024-04-27-DC-SLC")) |>
  # Here, it is the team in possession
  mutate(team = if_else(is_home_team == TRUE, home_teamID, away_teamID))
  # filter(team == "breeze")

test_data
```

```{r}
# Method 1 of calculating goals
test_data |>
  group_by(team, gameID) |>
  summarize(goal_scored_1 = max(away_team_score, na.rm = TRUE),
            .groups = "drop")
```

```{r}
# Method 2 of calculating goals
test_data |>
  group_by(team, gameID) |>
  summarize(goal_scored_2 = sum(goal == 1, na.rm = TRUE),
            .groups = "drop")
```
