---
format: gfm
---

## Case Study: Comparing attacking patterns of New York Empire and Tampa Bay Cannons

### I. DATA WRANGLING

#### 1. Read the team statistics data

The original CSV file is created using the `04_exploring_teams_data.qmd` file. Julia aggregated most of the team-specific data, and I added more metrics related to throwing patterns to this dataset.

```{r}
# Read the file and extract information of different teams
# (underperforming, mediocre, and elite teams)
rm(list = ls())
library(tidyverse)

full_team_stats <- read.csv("full_team_stats.csv", sep = ",")
full_team_stats
```

```{r}
# Elite teams (top 3): empire, shred, breeze. win_pct >= 0.8
# Mediocre teams (middle 3): radicals, growlers, glory. win_pct >= 0.5
# Underperforming teams (bottom 3): cannons, nitro, mechanix. win_pct lowest

selected_teams = c("empire", "shred", "breeze",
                   "radicals", "growlers", "glory",
                   "cannons", "nitro", "mechanix")

selected_team_stats <- full_team_stats |>
  filter(team %in% selected_teams) |>
  arrange(desc(win_pct))

selected_team_stats
```

#### 2. Read and add more metrics to the individual throw data

```{r}
ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")
ufa_throws
```

```{r}
ufa_throws_updated <- ufa_throws |>
  # Flipping the dataset so that positive x is the right side and negative x
  # is the left side
  mutate(thrower_x = (-1) * thrower_x,
         receiver_x = (-1) * receiver_x,
         throw_distance = sqrt((receiver_x-thrower_x)^2 + (receiver_y-thrower_y)^2),
         x_diff = receiver_x - thrower_x,
         throw_angle = atan2(y_diff, x_diff),
  # Add the teams in possession data
         team_in_possession = ifelse(is_home_team == TRUE,
                                     home_teamID, away_teamID),
  # Add the throw path type data for individual throws & attack outcome
         throw_path_type = case_when(
           (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
           (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
           TRUE ~ "cross_throws"
           ),
  # Add an identifier for each possession (attacking) phase in the dataset
  attacking_phaseID = consecutive_id(gameID, game_quarter, is_home_team,
                                     possession_num, home_team_score,
                                     away_team_score)) |>
  # Filter for selected teams
  filter(team_in_possession %in% selected_teams)

ufa_throws_updated
```

```{r}
# Add the attack outcome indicators (unique to each attacking phase ID)

# 1 = successful (leading to a goal)
# 0 = unsuccessful (ending in a turnover)
# -1 = incomplete (no goal nor turnover, likely because the quarter/match is ending)

attack_outcome_indicators <- ufa_throws_updated |>
  select(attacking_phaseID, goal, turnover) |>
  group_by(attacking_phaseID) |>
  summarize(sum_goal = sum(goal),
            sum_turnover = sum(turnover)) |>
  ungroup() |>
  mutate(attack_outcome = case_when(
    sum_goal == 1 ~ 1,
    sum_goal == 0 & sum_turnover >= 1 ~ 0,
    sum_goal == 0 & sum_turnover == 0 ~ -1
  )) |>
  select(attacking_phaseID, attack_outcome)

attack_outcome_indicators
```

```{r}
ufa_throws_updated <- ufa_throws_updated |>
  left_join(attack_outcome_indicators, join_by(attacking_phaseID))

ufa_throws_updated
```

### II. DATA ANALYSIS AND VISUALIZATION

#### 0. Define the preset themes for heat maps and clustered bar charts

```{r}
# https://stackoverflow.com/questions/23173915/can-ggplot-theme-formatting-be-saved-as-an-object

# Draw a circle on the plot with the geom_circle function from the ggforce package
# install.packages("ggforce")
library(ggforce)

# Set the classic theme
theme_set(theme_classic())

# Store the ggplot configurations for future use

#### 1. The upper half of a frisbee field ####

upper_half_frisbee_field <- list(
  
  # Limit the scale of variables shown in the plot
  # For shots, we only consider those made in the top half of the field
  # because shot data in other areas are so scarce
  # https://stackoverflow.com/questions/22945651/remove-space-between-plotted-data-and-the-axes
  scale_x_continuous(limits = c(-(26+2/3), 26+2/3), expand = c(0, 0),
                     breaks = seq(-20, 20, 10)),
  scale_y_continuous(limits = c(60, 120), expand = c(0, 0),
                     breaks = seq(60, 120, 10)),
  
  # Adding the frisbee field image to the plot
  # https://stackoverflow.com/questions/32226513/limiting-the-x-axis-range-of-geom-line-defined-by-slope-and-intercept
  # https://ggforce.data-imaginist.com/reference/geom_circle.html
  
  coord_fixed(),
  
  # I like the alternating texture of the grass field :)
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 110, ymax = 120, fill = "#008200"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 100, ymax = 110, fill = "#00aa00"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 90, ymax = 100, fill = "#008200"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 80, ymax = 90, fill = "#00aa00"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 70, ymax = 80, fill = "#008200"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 60, ymax = 70, fill = "#00aa00"),
  
  geom_segment(x = -(26+2/3), xend = 26+2/3,
               y = 100, yend = 100,
               color = "white", linewidth = 1.1),
  
  geom_circle(aes(x0 = 0, y0 = 110, r = 0.5), fill = "white", color = "white"),
  geom_circle(aes(x0 = 0, y0 = 80, r = 0.5), fill = "white", color = "white")
  
)

#### 2. The full frisbee field ####

lower_half_frisbee_field <- list(
  
  # Overwrite the existing scale of the upper half of the field
  scale_y_continuous(limits = c(0, 120), expand = c(0, 0),
                     breaks = seq(0, 120, 10)),
  
  # Draw the remaining grass texture to make a full frisbee field
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 50, ymax = 60, fill = "#008200"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 40, ymax = 50, fill = "#00aa00"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 30, ymax = 40, fill = "#008200"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 20, ymax = 30, fill = "#00aa00"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 10, ymax = 20, fill = "#008200"),
  
  annotate("rect", xmin = -(26+2/3), xmax = 26+2/3,
           ymin = 0, ymax = 10, fill = "#00aa00"),
  
  # Draw the midpoint line and own end line
  
  geom_segment(x = -(26+2/3), xend = 26+2/3,
               y = 60, yend = 60,
               color = "white", linewidth = 1.1),
  
  geom_segment(x = -(26+2/3), xend = 26+2/3,
               y = 20, yend = 20,
               color = "white", linewidth = 1.1),
  
  # Draw the circles for brick marks and reverse brick marks
  geom_circle(aes(x0 = 0, y0 = 40, r = 0.5), fill = "white", color = "white"),
  geom_circle(aes(x0 = 0, y0 = 10, r = 0.5), fill = "white", color = "white")
  
)

# Note that we are using the existing upper_half_frisbee_field list, which initially
# creates a list of lists.

# We need to apply the unlist() function to make it a 1D list to be passed
# as parameters in our ggplot call.
```


#### 1. Teams' attacking styles through heat maps

```{r}
ufa_throws_updated |>
  
  # Use the data for Empire and Cannons only, filter with the throws on goal
  filter(goal == 1, team_in_possession %in% c("empire", "cannons")) |>
  mutate(team_in_possession = ifelse(
    team_in_possession == "empire",
    "New York Empire (2021-24)", "Tampa Bay Cannons (2021-22)"
    )) |>
  
  ggplot() +
  
  upper_half_frisbee_field +
  
  # The main plot to create the heat map
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y), alpha = 0.8,
                         breaks = seq(0.0001, 0.0011, 0.0001)) +
  
  facet_wrap(~ team_in_possession, nrow = 1,
             axes = "all_y") +
  # axes = "all_y" to display the y-axis in both plots
  
  # Set the heat map's palette
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  # Note that we want to add "Direction = -1" so that Green represents the area
  # with the lowest density
  
  # Add labels to the plot
  labs(title = "Where points are scored",
       subtitle = "Throw locations of all assists for two Ultimate Frisbee teams",
       caption = "Data courtesy of Ultimate Frisbee Association & Eberhard, Miller, and Sandholtz (2025).",
       x = "Distance from the middle of the field in yards (positive = right side)",
       y = "Distance from backline at team's own end zone in yards") +
  
  # Get rid of gray background on the top of each facet plot (strip.background)
  # Remove gridlines from the plots (panel.grid)
  theme(strip.background = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "italic"),
        legend.position = "none")

# Save the resulting plot to display in the slides
ggsave("case_study_points_scored.png")
```

Locations of all successful throws (without a turnover, regardless of attacking phase) by team:

```{r}
successful_throws <- ufa_throws_updated |>
  
  filter(turnover == 0 & team_in_possession %in% c("empire", "cannons")) |>
  mutate(team_in_possession = ifelse(
    team_in_possession == "empire",
    "New York Empire (2021-24)", "Tampa Bay Cannons (2021-22)"
    ))

successful_throws
```


```{r}
# DO NOT RUN UNTIL I HAVE MADE ENOUGH CHANGES #

successful_throws |>
  
  ggplot() +
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y), alpha = 0.8) +
  facet_wrap(~ team_in_possession, nrow = 1, axes = "all_y") +
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  scale_y_continuous(limits = c(0, 120), expand = c(0, 0),
                     breaks = seq(0, 120, 10)) +
  
  #upper_half_frisbee_field + lower_half_frisbee_field +
  
  labs(title = "Where successful throws are made",
       subtitle = "Initial throw locations of all completed throws for two Ultimate Frisbee teams",
       caption = "Data courtesy of Ultimate Frisbee Association & Eberhard, Miller, and Sandholtz (2025).",
       x = "Distance from the middle of the field in yards (positive = right side)",
       y = "Distance from backline at team's own end zone in yards") +
  
  theme(strip.background = element_blank(),
        panel.grid = element_blank(),
        panel.spacing = unit(2, "lines"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "italic"),
        legend.position = "none")
```


Locations of turnovers by team:

```{r}
ufa_throws_updated |>
  
# Use the data for Empire and Cannons only, filter with the throws on goal
  filter(turnover == 1, team_in_possession %in% c("empire", "cannons")) |>
  mutate(team_in_possession = ifelse(
    team_in_possession == "empire",
    "New York Empire (2021-24)", "Tampa Bay Cannons (2021-22)"
    )) |>
  
  ggplot() +
  
  upper_half_frisbee_field + lower_half_frisbee_field +
  
  # The main plot to create the heat map
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y), alpha = 0.8) +
  
  facet_wrap(~ team_in_possession, nrow = 1,
             axes = "all_y") +
  # axes = "all_y" to display the y-axis in both plots
  
  # Set the heat map's palette
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  # Note that we want to add "Direction = -1" so that Green represents the area
  # with the lowest density
  
  # Add labels to the plot
  labs(title = "Where failed throws are made",
       subtitle = "Initial throw locations of all turnovers for two Ultimate Frisbee teams",
       caption = "Data courtesy of Ultimate Frisbee Association & Eberhard, Miller, and Sandholtz (2025).",
       x = "Distance from the middle of the field in yards (positive = right side)",
       y = "Distance from backline at team's own end zone in yards") +
  
  # Get rid of gray background on the top of each facet plot (strip.background)
  # Remove gridlines from the plots (panel.grid)
  
  # https://stackoverflow.com/questions/3681647/ggplot-how-to-increase-spacing-between-faceted-plots
  theme(strip.background = element_blank(),
        panel.grid = element_blank(),
        panel.spacing = unit(2, "lines"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "italic"),
        legend.position = "none")
```


Locations of throws (successful throws in a successful attacking phase) by team:

```{r}
ufa_throws_updated |>
  
# Use the data for Empire and Cannons only, filter with the throws on goal
  filter(attack_outcome == 1, team_in_possession %in% c("empire", "cannons")) |>
  mutate(team_in_possession = ifelse(
    team_in_possession == "empire",
    "New York Empire (2021-24)", "Tampa Bay Cannons (2021-22)"
    )) |>
  
  ggplot() +
  
  upper_half_frisbee_field + lower_half_frisbee_field +
  
  # The main plot to create the heat map
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y), alpha = 0.8) +
  
  facet_wrap(~ team_in_possession, nrow = 1,
             axes = "all_y") +
  # axes = "all_y" to display the y-axis in both plots
  
  # Set the heat map's palette
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  # Note that we want to add "Direction = -1" so that Green represents the area
  # with the lowest density
  
  # Add labels to the plot
  labs(title = "What are the successful locations?",
       subtitle = "Locations of successful throws in successful attacking phases for two Ultimate Frisbee teams",
       caption = "Data courtesy of Ultimate Frisbee Association & Eberhard, Miller, and Sandholtz (2025).",
       x = "Distance from the middle of the field in yards (positive = right side)",
       y = "Distance from backline at team's own end zone in yards") +
  
  # Get rid of gray background on the top of each facet plot (strip.background)
  # Remove gridlines from the plots (panel.grid)
  
  # https://stackoverflow.com/questions/3681647/ggplot-how-to-increase-spacing-between-faceted-plots
  theme(strip.background = element_blank(),
        panel.grid = element_blank(),
        panel.spacing = unit(2, "lines"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "italic"),
        legend.position = "none")
```


#### Bonus: Motivation for the case study

My main motivations to choose the New York Empire versus the Tampa Bay Cannons are not only their results (one has a very high and the other has a very low percentage).

The location of throws on goal ("ultimate frisbee's definition of assists") of the Empire and the Cannons look the same:

```{r}
# Use all the data for the top 3, middle 3, and bottom 3

ufa_throws_updated |>
  filter(goal == 1) |>
  ggplot() +
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y)) +
  scale_x_continuous(limits = c(-(26+2/3), 26+2/3)) +
  # For shots, we only consider those made after the midpoint
  # because shot data in other areas are so scarce
  scale_y_continuous(limits = c(60, 110)) +
  coord_fixed() + scale_fill_viridis_d() +
  geom_hline(yintercept = 100, color = "white") +
  theme(legend.position = "none") +
  facet_wrap(~ team_in_possession, nrow = 3, ncol = 3)
```

#### 2. Teams' statistics through numeric comparisons

```{r}
library(ggimage)

pairwise_comparison_stats <- full_team_stats |>
  filter(team %in% c("empire", "cannons")) |>
  add_column(logo_url = c("https://3.bp.blogspot.com/-ALvWHNs2RlU/Wp2bEgIcRgI/AAAAAAAAVb4/Uj4esHR2lEoLAmjE7p2xU4ov5mvf8ikHQCLcBGAs/s1600/tampa-bay-cannons_audl.png", "https://upload.wikimedia.org/wikipedia/en/2/27/New_York_Empire_%28AUDL%29_logo.png"))

pairwise_comparison_stats
```

```{r}
# https://stackoverflow.com/questions/20529252/change-or-modify-x-axis-tick-labels-in-r-using-ggplot2
# install.packages("ggimage")

pairwise_comparison_stats |>
  select(team, cross_throw_rate, left_wing_rate, right_wing_rate) |>
  pivot_longer(cross_throw_rate:right_wing_rate,
               names_to = "metric", values_to = "value") |>
  mutate(team = as.factor(team)) |>
  ggplot(aes(x = metric, y = value, fill = team)) +
  geom_col(position = "dodge", color = "gray40") +
  #geom_image(aes(image = logo_url)) +
  scale_x_discrete(labels = c("Cross throw rate", "Left-wing throw rate",
                              "Right-wing throw rate")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("#e6e600", "#007800"),
                    labels = c("Tampa Bay Cannons", "New York Empire")) +
  labs(title = "How often does an UFA/AUDL team throw in a certain direction?",
       subtitle = "Tampa Bay Cannons (2021-22) and New York Empire (2021-24)",
       caption = "Data courtesy of Ultimate Frisbee Association & Eberhard, Miller, and Sandholtz (2025).",
       x = "", y = "", fill = "Team") +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "italic"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

```{r}
pairwise_comparison_stats |>
  select(team, short_throw_rate, medium_throw_rate, long_throw_rate) |>
  pivot_longer(short_throw_rate:long_throw_rate,
               names_to = "metric", values_to = "value") |>
  mutate(team = as.factor(team)) |>
  ggplot(aes(x = metric, y = value, fill = team)) +
  geom_col(position = "dodge", color = "gray40") +
  scale_x_discrete(labels = c("Long throw rate", "Medium throw rate",
                              "Short throw rate")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("#e6e600", "#007800"),
                    labels = c("Tampa Bay Cannons", "New York Empire")) +
  labs(title = "How far does an UFA/AUDL team throw in a certain direction?",
       subtitle = "Tampa Bay Cannons (2021-22) and New York Empire (2021-24)",
       caption = "Data courtesy of Ultimate Frisbee Association & Eberhard, Miller, and Sandholtz (2025).",
       x = "", y = "", fill = "Team") +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "italic"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

```{r}
pairwise_comparison_stats |>
  select(team, avg_throw_distance, avg_goal_distance, avg_turnover_distance) |>
  pivot_longer(avg_throw_distance:avg_turnover_distance,
               names_to = "metric", values_to = "value") |>
  mutate(team = as.factor(team)) |>
  ggplot(aes(x = metric, y = value, fill = team)) +
  geom_col(position = "dodge", color = "gray40") +
  scale_x_discrete(labels = c("Average throw distance",
                              "Average scoring distance",
                              "Average turnover distance")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("#e6e600", "#007800"),
                    labels = c("Tampa Bay Cannons", "New York Empire")) +
  labs(title = "Comparing average offensive distance metrics of two UFA/AUDL teams",
       subtitle = "Tampa Bay Cannons (2021-22) and New York Empire (2021-24)",
       caption = "Data courtesy of Ultimate Frisbee Association & Eberhard, Miller, and Sandholtz (2025).",
       x = "", y = "Distance from own backline (yards)", fill = "Team") +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "italic"),
        axis.title.x = element_blank())
```
