---
format: gfm
---
## **EXPLANATORY DATA ANALYSIS (EDA) PROJECT: UFA THROWING STATISTICS**

## *Carnegie Mellon Sports Analytics Camp (CMSACamp) 2026*

### *An Le's Data Exploration*

#### 0. Importing the dataset

```{r}
rm(list = ls())

library(tidyverse)

ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")
head(ufa_throws)
```

#### 1. Exploring goal and turnover data

Relationship of each variable to whether a throw results in a goal:

```{r}
goal_throws <- ufa_throws |>
  filter(goal == 1)

failed_throws <- ufa_throws |>
  filter(goal == 0 & turnover == 1)

pass_throws <- ufa_throws |>
  filter(goal == 0 & turnover == 0)

goal_throws
failed_throws
pass_throws
```

Goal distance histogram (goal vs no goal):

```{r}
goal_throws |>
  ggplot(aes(x = throw_distance)) +
  geom_histogram()

failed_throws |>
  ggplot(aes(x = throw_distance)) +
  geom_histogram()

pass_throws |>
  ggplot(aes(x = throw_distance)) +
  geom_histogram()
```

```{r}
goal_throws |>
  ggplot(aes(x = throw_angle)) +
  geom_histogram() +
  coord_polar()

failed_throws |>
  ggplot(aes(x = throw_angle)) +
  geom_histogram() +
  coord_polar()

pass_throws |>
  ggplot(aes(x = throw_angle)) +
  geom_histogram() +
  coord_polar()
```

Coordinates of goal starting and ending points:

```{r}
# Scatter plot does not make too much sense
goal_throws |>
  ggplot(aes(x = thrower_x, y = thrower_y)) +
  geom_point() +
  coord_fixed() +
  scale_x_continuous(limits = c(min(ufa_throws$thrower_x), max(ufa_throws$thrower_x))) +
  scale_y_continuous(limits = c(0, 120))
```

```{r}
# What about box plots?
goal_throws |>
  ggplot(aes(x = thrower_x)) +
  geom_boxplot() +
  scale_x_continuous(limits = c(-30, 30))

goal_throws |>
  ggplot(aes(y = thrower_y)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 120))
```

How many throws lead to a goal?

```{r}
goal_throws |>
  ggplot(aes(x = possession_throw)) +
  stat_ecdf() +
  scale_x_continuous(breaks = seq(0, 65, 5))

# Around 70% of goals are created with <= 10 throws, and >50% of goals are created with <= 8 throws
```

Drawing arrows in ggplot2: `geom_segment`: https://r-graph-gallery.com/415-arrows-in-ggplot-graph.html

```{r}
goal_throws |>
  # Filter for a specific match for simplicity
  # Can uncomment the filter lines if needed
  filter(gameID == "2023-06-24-DC-BOS") |>
  ggplot() +
  geom_segment(aes(x = thrower_x, y = thrower_y,
                   xend = receiver_x, yend = receiver_y),
               arrow = arrow(), linewidth = 0.5) +
  coord_fixed()
```

What about a particular sequence of shots that lead to a goal?

```{r}
ufa_throws |>
  filter(gameID == "2023-06-24-DC-BOS" & times >= 2815) |>
  ggplot() +
  geom_segment(aes(x = thrower_x, y = thrower_y,
                   xend = receiver_x, yend = receiver_y),
               arrow = arrow(), linewidth = 0.5) +
  coord_fixed()
```

Try creating a heat map of position of goal throws by quarter:

```{r}
# https://sportscidata.com/2019/03/26/how-to-create-gps-heatmaps-in-ggplot/
# https://r-charts.com/correlation/contour-plot-ggplot2/

goal_throws |>
  ggplot(aes(x = thrower_x, y = thrower_y)) +
  geom_density_2d_filled() +
  scale_x_continuous(limits = c(-(26+2/3), 26+2/3)) +
  #scale_y_continuous(limits = c(0, 120)) +
  # Here we zoom in the "after-midpoint" position of the field
  scale_y_continuous(limits = c(60, 110)) +
  coord_fixed() + scale_fill_viridis_d() +
  theme(legend.position = "none") +
  geom_hline(yintercept = 100, color = "white") +
  facet_wrap(~ game_quarter, nrow = 2, ncol = 3)
```

Now do the same with the throws that make a turnover:

```{r}
failed_throws |>
  ggplot() +
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y)) +
  geom_point(aes(x = receiver_x, y = receiver_y), color = "red",
             alpha = 0.2, shape = 4, size = 0.5) +
  scale_x_continuous(limits = c(-(26+2/3), 26+2/3)) +
  scale_y_continuous(limits = c(0, 120)) +
  coord_fixed() + scale_fill_viridis_d() +
  theme(legend.position = "none") +
  geom_hline(yintercept = 20, color = "white") +
  geom_hline(yintercept = 100, color = "white") +
  facet_wrap(~ game_quarter, nrow = 1, ncol = 6)
```

Probability from scoring a goal and making a turnover from a certain coordinate along the y-axis:

```{r}
ufa_goal_turnover <- ufa_throws |>
  select(thrower_y, goal, turnover) |>
  mutate(zone = case_when(0 <= thrower_y & thrower_y <= 20 ~ "own_end_zone",
                          thrower_y > 20 & thrower_y <= 60 ~ "before_midpoint",
                          thrower_y > 60 ~ "after_midpoint"))

goal_from_zones <- ufa_goal_turnover |>
  select(zone, goal) |>
  group_by(zone) |>
  summarize(total_goals = sum(goal)) |>
  mutate(proportion = total_goals / sum(total_goals))

turnover_from_zones <- ufa_goal_turnover |>
  select(zone, turnover) |>
  group_by(zone) |>
  summarize(total_turnovers = sum(turnover)) |>
  mutate(proportion = total_turnovers / sum(total_turnovers))

goal_from_zones
turnover_from_zones
```

Throw distance based on the current team's lead:

```{r}
throw_dist_lead <- ufa_throws |>
  # Select both successful and unsuccessful throws
  select(throw_distance, is_home_team, score_diff) |>
  # Note that score_diff = score_home - score_away.
  # If the current team is not the home team, then we need to flip the signs
  mutate(adjusted_score_diff =
           case_when(is_home_team == TRUE ~ score_diff,
                     is_home_team == FALSE ~ (-1) * score_diff)) |>
  # Compute the median throwing distance with each score difference
  select(adjusted_score_diff, throw_distance) |>
  group_by(adjusted_score_diff) |>
  summarize(median_throw_distance = median(throw_distance))

throw_dist_lead

throw_dist_lead |>
  ggplot(aes(x = adjusted_score_diff, y = median_throw_distance)) +
  geom_point() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_smooth()
```

What if we only keep the successful throws?

```{r}
adjusted_throw_dist_lead <- ufa_throws |>
  filter(turnover == 0) |>
  select(throw_distance, is_home_team, score_diff) |>
  # Note that score_diff = score_home - score_away.
  # If the current team is not the home team, then we need to flip the signs
  mutate(adjusted_score_diff =
           case_when(is_home_team == TRUE ~ score_diff,
                     is_home_team == FALSE ~ (-1) * score_diff)) |>
  # Compute the median throwing distance with each score difference
  select(adjusted_score_diff, throw_distance) |>
  group_by(adjusted_score_diff) |>
  summarize(median_throw_distance = median(throw_distance))

adjusted_throw_dist_lead

adjusted_throw_dist_lead |>
  ggplot(aes(x = adjusted_score_diff, y = median_throw_distance)) +
  geom_point() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_smooth()
```

What if we adjust the dataset so that we are interested in how much higher up the field that the throw progressed?

```{r}
adjusted_throw_dist_lead_2 <- ufa_throws |>
  select(y_diff, is_home_team, score_diff) |>
  mutate(adjusted_score_diff =
           case_when(is_home_team == TRUE ~ score_diff,
                     is_home_team == FALSE ~ (-1) * score_diff)) |>
  # Compute the median throwing distance with each score difference
  select(adjusted_score_diff, y_diff) |>
  group_by(adjusted_score_diff) |>
  summarize(mean_y_diff = mean(y_diff))

adjusted_throw_dist_lead_2

adjusted_throw_dist_lead_2 |>
  ggplot(aes(x = adjusted_score_diff, y = mean_y_diff)) +
  geom_point() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_smooth()
```

Most of the failed passes (leading to the turnover) come before the midpoint, while the vast majority of the goal comes after the midpoint!

=\> How can teams break through the midpoint to score goals?

Note that field dimensions in ultimate frisbee may differ by tournaments and styles of play. In UFA, the field is 53+(1/3) yards wide and 80 yards long plus 20 yard endzones.

The UFA rulebook: https://watchufa.com/sites/default/files/UFA%20Rule%20Book%202024%20v12.0.pdf

#### 2. Exploring team results data

Which team appears the most in the dataset? (Removing entries for duplicated matches)

```{r}
# https://tidyr.tidyverse.org/reference/separate.html
# https://tidyr.tidyverse.org/reference/separate_wider_delim.html

# https://www.geeksforgeeks.org/r-language/concatenate-two-strings-in-r-programming-paste-method/

match_data <- ufa_throws |>
  select(gameID, home_teamID, away_teamID, home_team_win) |>
  distinct(gameID, .keep_all = TRUE) |>
  tidyr::separate_wider_delim(gameID, delim = "-",
                              names = c("game_year", "game_month", "game_date",
                                        NA, NA)) |>
  mutate(winner = case_when(home_team_win == 0 ~ away_teamID,
                            home_team_win == 1 ~ home_teamID),
         loser = case_when(winner == home_teamID ~ away_teamID,
                           winner == away_teamID ~ home_teamID),
         game_full_date = paste(game_year, game_month, game_date, sep = "-")) |>
  mutate(game_full_date = as.Date(game_full_date, "%Y-%m-%d")) |>
  select(game_full_date, home_teamID, away_teamID, winner, loser) |>
  arrange(game_full_date)

match_data
```

```{r}
summary(match_data$game_full_date)
```

June 4, 2021 is the first day of regular season of the 2021 season. August 24, 2024 is the championship game of the 2024 season.

https://www.watchufa.com/league/news/2021-audl-game-preview-week-1-minnesota-wind-chill-madison-radicals

https://www.watchufa.com/league/news/2024-ufa-playoffs-schedule

```{r}
home_appearances <- match_data |>
  count(home_teamID)

away_appearances <- match_data |>
  count(away_teamID)

# There is also an "allstar" game, which is useful to remove in this case
total_appearances <- home_appearances |>
  full_join(away_appearances, join_by(home_teamID == away_teamID),
            suffix = c("_home", "_away")) |>
  mutate(total_matches = n_home + n_away) |>
  rename(teamID = home_teamID) |>
  filter(!(teamID %in% c("allstars1", "allstars2")))

total_appearances
```

```{r}
# Pivoting and plotting as a facet
total_appearances_pivot <- total_appearances |>
  pivot_longer(n_home:total_matches,
               names_to = "match_type", values_to = "count") |>
  filter(match_type != "total_matches")

total_appearances_pivot

total_appearances_pivot |>
  ggplot(aes(x = teamID, y = count, fill = match_type)) +
  geom_col(position = "dodge") +
  # Change position of legend to bottom, rotate x-axis labels by 45 degrees
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
  #facet_grid(match_type ~ .)
```

What about the number of wins or losses?

```{r}
win_counts <- match_data |>
  count(winner)

loss_counts <- match_data |>
  count(loser)

# There is also an "allstar" game, which is useful to remove in this case
overall_stats <- win_counts |>
  left_join(loss_counts, join_by(winner == loser),
            suffix = c("_wins", "_losses")) |>
  mutate(total_matches = n_wins + n_losses) |>
  rename(teamID = winner) |>
  filter(!(teamID %in% c("allstars1", "allstars2")))

overall_stats
```

```{r}
# Same as the previous use case
overall_stats_pivot <- overall_stats |>
  pivot_longer(n_wins:total_matches,
               names_to = "outcome", values_to = "count") |>
  filter(outcome != "total_matches")

overall_stats_pivot

overall_stats_pivot |>
  ggplot(aes(x = teamID, y = count, fill = outcome)) +
  geom_col(position = "dodge") +
  # Change position of legend to bottom, rotate x-axis labels by 45 degrees
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
  #facet_grid(match_type ~ .)
```

```{r}
# How about the win percentage?
overall_stats_win_pct <- overall_stats |>
  mutate(win_percentage = n_wins / total_matches,
# Use example from Lecture 3, mutate to display it properly in the following plot
         teamID = fct_reorder(teamID, win_percentage, .desc = TRUE)) |>
  arrange(desc(win_percentage))

overall_stats_win_pct

overall_stats_win_pct |>
  ggplot(aes(x = teamID, y = win_percentage)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

#### 3. Proposed questions

1.  Which direction of attack (i.e. to the right or left side of the field) has more chances of scoring a goal? Does this vary across teams?

2.  What makes a team attempt further throws, or attempt to score goals with fewer passes? Is it the time pressure (little time left in quarter or the match), or the score pressure (being led by so many points)?

3.  For all teams' possession period that builds up from before the midpoint and passing through the midpoint, how likely is a team to score goals afterwards?

4.  Explore questions on how a particular team's throwing strategies evolved over time, or how a particular tournament-wide statistic has changed over time:

- Number of throws that make a goal / throw distance of a goal / total time that it takes for teams to score goals
- Whether teams attack more on the left wing or on the right wing
- Locations of turnovers (how far has a team gone up the field when they made a turnover?)

"Over time" can mean comparing between quarters, between different seasons, or between regular season and playoffs (need additional research and data processing).

5.  Clustering analysis: Cluster teams by the attacking strategy (starting vertical half, ending vertical half, number of times that the team switch vertical halves).

6.  Can also repeat these questions but for a particular team (eg: Pittsburgh Thunderbirds) to analyze in depth and suggest strategies.
