---
format: gfm
---
#### 0. Importing the dataset

```{r}
rm(list = ls())

library(tidyverse)

ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")

ufa_throws
```

```{r}
# Flipping the dataset so that positive x is the right side and negative x is the left side
ufa_throws <- ufa_throws |>
  mutate(thrower_x = (-1) * thrower_x,
         receiver_x = (-1) * receiver_x,
         throw_distance = sqrt((receiver_x-thrower_x)^2 + (receiver_y-thrower_y)^2),
         x_diff = receiver_x - thrower_x,
         throw_angle = atan2(y_diff, x_diff))

ufa_throws
```

```{r}
# Filtering for successful throws
successful_throws <- ufa_throws |>
  filter(turnover == 0)

successful_throws
```

#### Main question:

**Which direction of attack (i.e. to the right or left side of the field) has more chances of scoring a goal? Does this vary across teams?**

See the `01_data_exploration.qmd` files for:

- Heat map for the directions of scoring a goal (faceted by quarter)
- Example of a sequence of throws that lead to a goal

Note: A sequence of throws eventually leading to a turnover/goal is (hopefully) uniquely determined by the following variables:

- Possession number
- Game quarter
- Team in possession (`if_home_team`)
- Current scores of home and away teams
- Game ID

#### 1. Number of throws made in a particular wing of the field or crossed the midfield (successful throws only)

```{r}
successful_throws |>
  mutate(throw_path_type = case_when(
    (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
    (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
    (thrower_x >= 0) & (receiver_x <= 0) ~ "right_to_left",
    (thrower_x <= 0) & (receiver_x >= 0) ~ "left_to_right"
  )) |>
  count(throw_path_type) |>
  mutate(prop = n / sum(n))
```

#### 1.1. Additional filter step for successful attacking phases (those that result in a goal)

```{r}
# Create a list of attacking phases that result in a goal
successful_goals_indicators <- ufa_throws |>
  filter(goal == 1) |>
  select(gameID, game_quarter, is_home_team, possession_num,
         home_team_score, away_team_score)

successful_goals_indicators
```

```{r}
# Now only keep these attacking phases in the UFA throws table
successful_attacking_phases <- successful_goals_indicators |>
  left_join(ufa_throws,
            join_by(gameID, game_quarter, is_home_team, possession_num,
                    home_team_score, away_team_score))

successful_attacking_phases
```

#### 2. Number of left-wing, right-wing, and cross-wing passes

Overall distribution across the dataset:

```{r}
successful_attacking_phases <- successful_attacking_phases |>
  mutate(throw_path_type = case_when(
    (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
    (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
    TRUE ~ "cross_throw"
    ),
    team_in_possession = case_when(
      is_home_team == TRUE ~ home_teamID,
      is_home_team == FALSE ~ away_teamID
    ))

successful_attacking_phases
```

```{r}
count_throw_path_type <- successful_attacking_phases |>
  group_by(gameID, game_quarter, team_in_possession, possession_num,
           home_team_score, away_team_score) |>
  # Count the throw path type (which wing?) by each attacking phase in all matches
  count(throw_path_type) |>
  ungroup()

count_throw_path_type
```

```{r}
# Note that we are working on successful attacking phases: the attacks that eventually lead to a goal

throw_path_by_team <- count_throw_path_type |>
  select(team_in_possession, throw_path_type, n) |>
  # Count of different types of throws by team
  group_by(team_in_possession, throw_path_type) |>
  summarize(count_throw_path = n()) |>
  ungroup()

throw_path_by_team

sum_throws_by_team <- count_throw_path_type |>
  select(team_in_possession, n) |>
  # Total of throws by team in possession
  count(team_in_possession)

sum_throws_by_team

prop_throws_by_team <- throw_path_by_team |>
  left_join(sum_throws_by_team, join_by(team_in_possession)) |>
  mutate(prop = count_throw_path / n) |>
  filter(!(team_in_possession %in% c("allstars1", "allstars2")))

prop_throws_by_team
```

```{r}
# Which teams have the highest proportion of cross throws?

prop_throws_by_team |>
  filter(throw_path_type == "cross_throw") |>
  select(team = team_in_possession, prop) |>
  ggplot(aes(y = team, x = prop)) +
  geom_col()
```

Number of cross-wing shots (last passes):

```{r}
successful_attacking_phases |>
  filter(goal == 1) |>
  count(throw_path_type)
```

Number of goals that a team has made after picking up from at least a turnover:

```{r}
successful_attacking_phases |>
  filter(possession_num > 1 & goal == 1) |>
  count(possession_num)
  #ggplot(aes(x = possession_num, y = n)) +
  #geom_col()
```

There are actually a lot of them! (`4416 / 21391 = approx. 20.6%`)

#### 3. Which place up the field (y-coordinate) are teams most likely to change their wing of attack?

```{r}
# Here we consider successful passes (no turnovers)

successful_throw_types <- successful_throws |>
  mutate(throw_path_type = case_when(
    (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
    (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
    TRUE ~ "cross_throw"
    ))

successful_throw_types

successful_throw_types |>
  filter(throw_path_type == "cross_throw") |>
  ggplot(aes(y = thrower_y)) +
  geom_histogram()
```

```{r}
# What if we plot them as positional heat map data?
# Much of the settings are taken from my `01_data_exploration.qmd` file

successful_throw_types |>
  # We also need the pass to have a considerable distance into each wing
  # to be considered a cross
  filter(throw_path_type == "cross_throw" &
           (abs(thrower_x) >= 5 | abs(receiver_x) >= 5)) |>
  ggplot() +
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y)) +
  scale_x_continuous(limits = c(-(26+2/3), 26+2/3)) +
  scale_y_continuous(limits = c(0, 120)) +
  coord_fixed() + scale_fill_viridis_d() +
  geom_hline(yintercept = 20, color = "white") +
  geom_hline(yintercept = 100, color = "white") +
  theme(legend.position = "none")
```

Much of the activity is happening before the midpoint (60-yard line), or more precisely, around 25 to 50 yards up the field.

What if we facet this plot by another variable, such as whether the attacking phase is successful or not?

I will denote this using a binary variable called `attack_outcome`, which takes 1 if attacking phase eventually results in a goal, and 0 if resulting in a turnover.

```{r}
# Create another dataset of unsuccessful attacking phases
# Similar to the successful attacking phases data outlined earlier in this notebook
turnover_indicators <- ufa_throws |>
  filter(turnover == 1) |>
  select(gameID, game_quarter, is_home_team, possession_num,
         home_team_score, away_team_score)

turnover_indicators

unsuccessful_attacking_phases <- turnover_indicators |>
  left_join(ufa_throws,
            join_by(gameID, game_quarter, is_home_team, possession_num,
                    home_team_score, away_team_score)) |>
  mutate(attack_outcome = 0,
         throw_path_type = case_when(
           (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
           (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
           TRUE ~ "cross_throw"
           ))

unsuccessful_attacking_phases
```

```{r}
successful_attacking_phases <- successful_attacking_phases |>
  mutate(attack_outcome = 1)

ufa_throws_updated <- bind_rows(successful_attacking_phases,
                                unsuccessful_attacking_phases)

ufa_throws_updated
```

```{r}
# Caveat: Not every attacking phase ends with a goal or with a turnover.
# We will look at this later
ufa_throws |>
  anti_join(ufa_throws_updated)
```

```{r}
ufa_throws_updated |>
  filter(throw_path_type == "cross_throw" &
           (abs(thrower_x) >= 5 | abs(receiver_x) >= 5)) |>
  ggplot() +
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y)) +
  facet_wrap(~ attack_outcome) +
  scale_x_continuous(limits = c(-(26+2/3), 26+2/3)) +
  scale_y_continuous(limits = c(0, 120)) +
  coord_fixed() + scale_fill_viridis_d() +
  geom_hline(yintercept = 20, color = "white") +
  geom_hline(yintercept = 100, color = "white") +
  theme(legend.position = "none")
```

Visualizing position of cross-throwing shots (that actually results in a goal):

```{r}
successful_attacking_phases |>
  # Here, we change the | into &
  # We want both the thrower and receiver to be at least 5 yards away from the
  # vertical center of the field to be considered as a cross shot on goal
  filter(goal == 1 & throw_path_type == "cross_throw" &
         abs(thrower_x) >= 5 & abs(receiver_x) >= 5) |>
  ggplot() +
  geom_density_2d_filled(aes(x = thrower_x, y = thrower_y)) +
  scale_x_continuous(limits = c(-(26+2/3), 26+2/3)) +
  # For shots, we only consider those made after the midpoint
  # because shot data in other areas are so scarce
  scale_y_continuous(limits = c(60, 110)) +
  coord_fixed() + scale_fill_viridis_d() +
  geom_hline(yintercept = 100, color = "white") +
  theme(legend.position = "none")
```

Most cross shots are made within the 80-85 yards mark.

How about adding another hexagon density plot version from Lecture 5?

```{r}
# install.packages("hexbin")
library(hexbin)

successful_attacking_phases |>
  filter(goal == 1 & throw_path_type == "cross_throw" &
         abs(thrower_x) >= 5 & abs(receiver_x) >= 5) |>
  ggplot(aes(x = thrower_x, y = thrower_y)) +
  geom_hex(binwidth = c(5, 5)) +
  scale_x_continuous(limits = c(-(26+2/3), 26+2/3)) +
  scale_y_continuous(limits = c(60, 110)) +
  coord_fixed() +
  geom_hline(yintercept = 100, color = "white") +
  scale_fill_gradient(low = "midnightblue", high = "goldenrod") +
  theme(legend.position = "none")
```

#### 3. Trends in left-wing, right-wing, and cross-wing by season

Here, we are taking statistics for all teams (also including all-star teams).

```{r}
ufa_throws_seasons <- ufa_throws |>
  mutate(season = as.numeric(substr(gameID, 1, 4)),
         throw_path_type = case_when(
           (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
           (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
           TRUE ~ "cross_throw"
           ))
  
ufa_throw_path_by_season <- ufa_throws_seasons |>
  select(season, throw_path_type) |>
  # Count the number of each throw path types by season
  group_by(season) |>
  count(throw_path_type) |>
  ungroup()

ufa_throw_path_by_season

ufa_throws_games_by_season <- ufa_throws_seasons |>
  select(season, gameID) |>
  unique() |>
  # Count the number of game IDs by season
  count(season)

ufa_throws_games_by_season
```

```{r}
ufa_throw_path_by_season_2 <- ufa_throw_path_by_season |>
  left_join(ufa_throws_games_by_season, join_by(season),
            suffix = c("_throws", "_match_in_season")) |>
  mutate(avg_throws_per_match = n_throws / n_match_in_season)

ufa_throw_path_by_season_2

ufa_throw_path_by_season_2 |>
  ggplot(aes(x = season, y = avg_throws_per_match, color = throw_path_type)) +
  geom_point() + geom_line() +
  scale_fill_viridis_d() + theme(legend.position = "bottom")
```

#### Other questions (will explore later):

- Interaction between the number of turnovers it takes a team to score goals and the goal differences / final scores of the match

Metric: Arranging teams by the number of turnovers or possession number before scoring a goal. Higher possession number = wasting many scoring chances.

- Explore the starting position (where to build up) and ending position (where there is a shot on goal) of a successful/unsuccessful attacking phase

I have partially explored that in the notebook, but there might be more variables to consider.

- Explore the possession time of a successful attacking phase, although this is outside of the throw position data focus of this notebook

- Trends in right/left/cross wing tendencies over time of the teams by season

- Tendency to attack right wing or left wing by team, such as proportion of primarily right/left wing attacking phases divided by the total number of attacking phases (can be included in the clustering analysis)

Caveat in ultimate frisbee data: Different from soccer where players need to move the ball to the midfield some time for a good shot on goal. In frisbee, players can build up their attack and score from anywhere.