## **EXPLANATORY DATA ANALYSIS (EDA) PROJECT: UFA THROWING STATISTICS**

## *Carnegie Mellon Sports Analytics Camp (CMSACamp) 2026*

### *An Le's Data Exploration*

#### 0. Importing the dataset

```{r}
rm(list = ls())

library(tidyverse)

ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")
head(ufa_throws)
```

#### 1. Exploring goal and turnover data

Relationship of each variable to whether a throw results in a goal:

```{r}
goal_throws <- ufa_throws |>
  filter(goal == 1)

failed_throws <- ufa_throws |>
  filter(goal == 0 & turnover == 1)

pass_throws <- ufa_throws |>
  filter(goal == 0 & turnover == 0)

goal_throws
failed_throws
pass_throws
```

Goal distance histogram (goal vs no goal):

```{r}
goal_throws |>
  ggplot(aes(x = throw_distance)) +
  geom_histogram()

failed_throws |>
  ggplot(aes(x = throw_distance)) +
  geom_histogram()

pass_throws |>
  ggplot(aes(x = throw_distance)) +
  geom_histogram()
```

```{r}
goal_throws |>
  ggplot(aes(x = throw_angle)) +
  geom_histogram() +
  coord_polar()

failed_throws |>
  ggplot(aes(x = throw_angle)) +
  geom_histogram() +
  coord_polar()

pass_throws |>
  ggplot(aes(x = throw_angle)) +
  geom_histogram() +
  coord_polar()
```

Coordinates of goal starting and ending points:

```{r}
# Scatter plot does not make too much sense
goal_throws |>
  ggplot(aes(x = thrower_x, y = thrower_y)) +
  geom_point() +
  coord_fixed() +
  scale_x_continuous(limits = c(min(ufa_throws$thrower_x), max(ufa_throws$thrower_x))) +
  scale_y_continuous(limits = c(0, 120))
```

```{r}
# What about box plots?
goal_throws |>
  ggplot(aes(x = thrower_x)) +
  geom_boxplot() +
  scale_x_continuous(limits = c(-30, 30))

goal_throws |>
  ggplot(aes(y = thrower_y)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 120))
```

How many throws lead to a goal?

```{r}
goal_throws |>
  ggplot(aes(x = possession_throw)) +
  stat_ecdf() +
  scale_x_continuous(breaks = seq(0, 65, 5))

# Around 70% of goals are created with <= 10 throws, and >50% of goals are created with <= 8 throws
```

Drawing arrows in ggplot2: `geom_segment`: https://r-graph-gallery.com/415-arrows-in-ggplot-graph.html

```{r}
goal_throws |>
  # Filter for a specific match for simplicity
  # Can uncomment the filter lines if needed
  filter(gameID == "2023-06-24-DC-BOS") |>
  ggplot() +
  geom_segment(aes(x = thrower_x, y = thrower_y,
                   xend = receiver_x, yend = receiver_y),
               arrow = arrow(), linewidth = 0.5) +
  coord_fixed()
```

Probability from scoring a goal and making a turnover from a certain coordinate along the y-axis:

```{r}
ufa_goal_turnover <- ufa_throws |>
  select(thrower_y, goal, turnover) |>
  mutate(zone = case_when(0 <= thrower_y & thrower_y <= 20 ~ "own_end_zone",
                          thrower_y > 20 & thrower_y <= 60 ~ "before_midpoint",
                          thrower_y > 60 ~ "after_midpoint"))

goal_from_zones <- ufa_goal_turnover |>
  select(zone, goal) |>
  group_by(zone) |>
  summarize(total_goals = sum(goal)) |>
  mutate(proportion = total_goals / sum(total_goals))

turnover_from_zones <- ufa_goal_turnover |>
  select(zone, turnover) |>
  group_by(zone) |>
  summarize(total_turnovers = sum(turnover)) |>
  mutate(proportion = total_turnovers / sum(total_turnovers))

goal_from_zones
turnover_from_zones
```

Most of the failed passes (leading to the turnover) come before the midpoint, while the vast majority of the goal comes after the midpoint!

=> How can teams break through the midpoint to score goals?

Note that field dimensions in ultimate frisbee may differ by tournaments and styles of play. In UFA, the field is 53+(1/3) yards wide and 80 yards long plus 20 yard endzones.

The UFA rulebook:
https://watchufa.com/sites/default/files/UFA%20Rule%20Book%202024%20v12.0.pdf

#### 2. Exploring team results data

Which team appears the most in the dataset? (Removing entries for duplicated matches)

```{r}
# https://tidyr.tidyverse.org/reference/separate.html
# https://tidyr.tidyverse.org/reference/separate_wider_delim.html

# https://www.geeksforgeeks.org/r-language/concatenate-two-strings-in-r-programming-paste-method/

match_data <- ufa_throws |>
  select(gameID, home_teamID, away_teamID, home_team_win) |>
  distinct(gameID, .keep_all = TRUE) |>
  tidyr::separate_wider_delim(gameID, delim = "-",
                              names = c("game_year", "game_month", "game_date",
                                        NA, NA)) |>
  mutate(winner = case_when(home_team_win == 0 ~ away_teamID,
                            home_team_win == 1 ~ home_teamID),
         loser = case_when(winner == home_teamID ~ away_teamID,
                           winner == away_teamID ~ home_teamID),
         game_full_date = paste(game_year, game_month, game_date, sep = "-")) |>
  mutate(game_full_date = as.Date(game_full_date, "%Y-%m-%d")) |>
  select(game_full_date, home_teamID, away_teamID, winner, loser) |>
  arrange(game_full_date)

match_data
```

```{r}
summary(match_data$game_full_date)
```

June 4, 2021 is the first day of regular season of the 2021 season.
August 24, 2024 is the championship game of the 2024 season.

https://www.watchufa.com/league/news/2021-audl-game-preview-week-1-minnesota-wind-chill-madison-radicals

https://www.watchufa.com/league/news/2024-ufa-playoffs-schedule

```{r}
home_appearances <- match_data |>
  count(home_teamID)

away_appearances <- match_data |>
  count(away_teamID)

# There is also an "allstar" game, which is useful to remove in this case
total_appearances <- home_appearances |>
  full_join(away_appearances, join_by(home_teamID == away_teamID),
            suffix = c("_home", "_away")) |>
  mutate(total_matches = n_home + n_away) |>
  rename(teamID = home_teamID) |>
  filter(!(teamID %in% c("allstars1", "allstars2")))

total_appearances
```

```{r}
# Pivoting and plotting as a facet
total_appearances_pivot <- total_appearances |>
  pivot_longer(n_home:total_matches,
               names_to = "match_type", values_to = "count") |>
  filter(match_type != "total_matches")

total_appearances_pivot

total_appearances_pivot |>
  ggplot(aes(x = teamID, y = count, fill = match_type)) +
  geom_col(position = "dodge") +
  # Change position of legend to bottom, rotate x-axis labels by 45 degrees
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
  #facet_grid(match_type ~ .)
```

What about the number of wins or losses?

```{r}
win_counts <- match_data |>
  count(winner)

loss_counts <- match_data |>
  count(loser)

# There is also an "allstar" game, which is useful to remove in this case
overall_stats <- win_counts |>
  left_join(loss_counts, join_by(winner == loser),
            suffix = c("_wins", "_losses")) |>
  mutate(total_matches = n_wins + n_losses) |>
  rename(teamID = winner) |>
  filter(!(teamID %in% c("allstars1", "allstars2")))

overall_stats
```

```{r}
# Same as the previous use case
overall_stats_pivot <- overall_stats |>
  pivot_longer(n_wins:total_matches,
               names_to = "outcome", values_to = "count") |>
  filter(outcome != "total_matches")

overall_stats_pivot

overall_stats_pivot |>
  ggplot(aes(x = teamID, y = count, fill = outcome)) +
  geom_col(position = "dodge") +
  # Change position of legend to bottom, rotate x-axis labels by 45 degrees
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
  #facet_grid(match_type ~ .)
```

```{r}
# How about the win percentage?
overall_stats_win_pct <- overall_stats |>
  mutate(win_percentage = n_wins / total_matches,
# Use example from Lecture 3, mutate to display it properly in the following plot
         teamID = fct_reorder(teamID, win_percentage, .desc = TRUE)) |>
  arrange(desc(win_percentage))

overall_stats_win_pct

overall_stats_win_pct |>
  ggplot(aes(x = teamID, y = win_percentage)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

Does a team perform better in certain times of the year or certain periods of the season? [TO BE UPDATED]

#### 3. Proposed questions

1.  Which direction of attack (i.e. to the right or left side of the field) has more chances of scoring a goal? Does this vary across teams?

2.  What makes a team attempt further throws, or attempt to score goals with fewer passes? Is it the time pressure (little time left in quarter or the match), or the score pressure (being led by so many points)?

3.  For all team's possession period that builds up from before the midpoint and passing through the midpoint, how likely is a team to score goals afterwards?

4.  [TO BE UPDATED] Can explore questions on how a particular team's throwing strategies evolved over time, or how a particular tournament-wide statistic has changed over time.

5.  Clustering analysis: Cluster teams by the attacking strategy (starting vertical half, ending vertical half, number of times that the team switch vertical halves).

6.  Can also repeat these questions but for a particular team (eg: Pittsburgh Thunderbirds) to analyze in depth and suggest strategies.
