---
title: "test-du"
format: html
---

```{r}
library(tidyverse)
library(lubridate)
ufa_throws <- read_csv("https://raw.githubusercontent.com/BradenEberhard/Expected-Throwing-Value/refs/heads/main/data/all_games_1024.csv")

ufa_throws <- ufa_throws |> 
  mutate(
    goal = ifelse(receiver_y > 100 & turnover == 0, 1, 0),
    throw_distance = sqrt((receiver_x - thrower_x) ^ 2 + (receiver_y - thrower_y) ^ 2),
    x_diff = receiver_x - thrower_x,
    y_diff = receiver_y - thrower_y,
    throw_angle = atan2(y_diff, x_diff)
  ) |> 
  select(-quarter_point, -total_points, -current_line)

ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")
```

# EDA 

## Game
gameID, possession_num, possession_throw, game_quarter, home_team_score, away_team_score, home_teamID, away_teamID, home_team_win (whether home team wins), score_diff (between home and away teams)

```{r}
ufa_throws |>
  filter(goal == 0 & turnover == 1) # intermediate passing throw, neither turnover nor goal (when turnover = 0, a throw could result in either a goal or a regular passing. When turnover = 1, this means the throw could not result in a goal)

thrower_turnover <- ufa_throws |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  group_by(year, thrower) |>
  summarize(num_turnover_in_a_season = sum(turnover)) |>
  ungroup() 

thrower_game <- ufa_throws|>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  group_by(year, thrower) |>
  summarize(num_game_in_a_season = n_distinct(gameID)) |>
  ungroup()

left_join(thrower_turnover, thrower_game, by = c("thrower", "year")) |>
  mutate(avg_turnover_per_game = num_turnover_in_a_season/num_game_in_a_season)
  
thrower_turnover |>
  ggplot(aes(x = avg_turnover_per_game)) +
  geom_histogram() + 
  labs(title = "Distribution of Average Thrower Turnover per Game")
```

```{r}
ufa_throws <- ufa_throws |>
  left_join(thrower_turnover, by = "thrower") |>
  relocate(avg_turnover_per_game)

ufa_throws |>
  group_by(home_teamID) |>
  arrange(desc(avg_turnover_per_game), .by_group = TRUE) |>
  ggplot(aes(x = ))
```

## Individual throwers/receivers
thrower, thrower_x, thrower_y, receiver, receiver_x, receiver_y

## Throw (results in goal/turnover, angle)
turnover, game_quarter, is_home_team, times (game time remaining in seconds), throw_distance, x_diff, y_diff, throw_angle 